---
title: "Bayesian Differential Expression Analysis for Proteomics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Differential Expression Analysis for Proteomics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
# Installation

```


# Main functions
- `BayesMissingModel`: Main function for doing differential expression analysis on data frame/matrix objects by calculating different         statistics of posterior distribution.

- `BayesianMissingAggregate`: Similar with `BayesMissingModel`, but for QFeatures objects.

- `plotPost`: Visualize the log fold change posterior distribution and relative statistics of a specified protein.


```{r setup}
library(missBayes)
library(limma)
```

# Example: Proteomic Analysis of FFPE Tissue Samples Identifies Potential Molecular Mechanisms Mediating Resistance to Radiotherapy in Rectal Cancer

This study aims to investigate pathophysiological mechanisms underlying radioresistance.Out of a total of 50 tumor patients, paired tissue (control and tumor) samples were available from 39 patients. 

Data is available from PRIDE with the id of PXD060201. The input data is MaxQuant output file proteinGroups, only a subset is used for example.


```{r}

# data pre-processing
data_path <- system.file("extdata", "PXD060201_vignettes.csv", package = "missBayes")
pg <- read.csv(data_path, blank.lines.skip = TRUE)
# select tumor and control samples 
pg <- pg[, c(1,558:646)]
pg <- pg[, c(-12,-15, -18, -19, -30, -31, -46, -47, -84, -85, -88)]

# set row names to protein ID and remove ID column
pg <- pg[pg$Protein.IDs != "" & !is.na(pg$Protein.IDs), ]
rownames(pg) <- pg$Protein.IDs
pg <- pg[, -1]

# log2 transformation, convert -Inf to NA and median normalization
log2.df <- log2(pg)
log2.df[is.infinite(as.matrix(log2.df))] <- NA
log2norm.df <- normalizeMedianValues(log2.df)

# filter out the rows that are entirely missing
log2norm.filtered.df <- log2norm.df[rowSums(!is.na(log2norm.df)) > 0, ]

# create group factor
group_factor <- ifelse(grepl("_Tu", colnames(log2norm.df)), "Tu",
                ifelse(grepl("_con", colnames(log2norm.df)), "con",
                NA))
group_factor <- as.factor(group_factor)

# create contrast matrix
comparison <- makeContrasts(Tu - con, levels = levels(group_factor))

```

```{r}
# run Bayesian missing model
output <- BayesMissingModel(log2norm.filtered.df, group_factor, comparison, parallel = FALSE)

```

```{r}
# view the logFC distribution of protein "Q96D15"
my_plot <- plotPost(log2norm.filtered.df, "Q96D15",group_factor, "Tu - con")


```
#References: 
Proteomic Analysis of FFPE Tissue Samples Identifies Potential Molecular Mechanisms Mediating Resistance to Radiotherapy in Rectal Cancer
Tobias Zott, Michael Wolf, Günter Plessl-Walder, Heinz Regele, Michael Bergmann, Samuel M. Meier-Menches, Christopher Gerner, Gerd R. Silberhumer, and Andrea Bileck
Journal of Proteome Research 2025 24 (8), 3990-4001
